---
title: "project_one"
author:
  - "Sanna Askman"
  - "Jonas Engström"
  - "Hanna Gustafsson"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{project_one}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Project 1

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8.9,
  fig.height = 5,
  out.width = "100%"
)
```

The instructions below are copied from [the Lund University website](https://canvas.education.lu.se/courses/34580/files/6061921?module_item_id=1441780).

## Prognostic factors in breast cancer---focus on uPA

### Background and Material

During the first half of the 1990s, 237 women with primary breast cancer, diagnosed in the South Swedish health care region, were included in cohort study. The primary aim was validation of a prognostic index based on tumor size, proliferation, and receptor status and the secondary to identify new biomarkers. Women with other known malignancies before the first breast cancer diagnosis were excluded. Furthermore, the study was restricted to pre-menopausal women with localized disease, i.e., no spread to the axillary lymph nodes. Fresh frozen tumor tissue was sent to the laboratory at the Department of Oncology, Lund, for routine evaluation of tumor characteristics such as proliferation (S-phase fraction) and receptor status (ER and PgR). Small pieces from these tumors have since then been used in numerous biomarker studies and we still have tissue left from most of these tumors in the biobank at Medicon Village, Lund.

The study database currently contains 124 variables and the number of variables is growing as new biomarkers are measured. The dataset you are going to analyze is a random sample of 200 patients and 22 variables from this study. Some of the variables are original and some derived. For example, the variable hist_grade_3_to_9 is the sum of the three variables mitotic_grade, tubular_grade and nuclear_grade. The histological grade of the tumour (hist_grade) is defined by categorization of this sum into three classes: Grade 1 (sum 3 to 5), Grade 2 (sum 6 or 7) and Grade 3 (sum 8 or 9). This variable has also been dichotomized, contrasting Grade 3 to Grade 1+2 (hist3).

Variables in the data file:

|variable name|storage type|display format|value label|variable label|
|-|-|-|-|-|
|id|float|%9.0g||Patient ID|
|age|float|%9.0g||age, years|
|adjuvant|byte|%9.0g|janej|Adjuvant treatment, hormonal or chemotherapy|
|ddfs|float|%9.0g||Distant disease-free survival, years|
|distmet|byte|%9.0g|janej|Distant metastasis?|
|tumsize|byte|%9.0g||Tumor size, mm|
|size_gt_20|byte|%9.0g|janej|Tumor size >20 mm?|
|hist_grade|byte|%4.0g||Histological grade|
|hist3|byte|%9.0g|janej|Histological grade 3?|
|mitotic_grade|byte|%4.0g||Mitotic grade|
|tubular_grade|byte|%4.0g||Tubular grade|
|nuclear_grade|byte|%4.0g||Nuclear grade|
|hist_grade_3_~9|byte|%9.0g||Sum of mitotic, tubular and nuclear grade|
|er_cyt|float|%9.0g||ER, fmol/mg protein cytosol|
|er_cyt_pos|byte|%9.0g|janej|ER cytosol>=25fmol/mg protein?|
|pgr_cyt|float|%9.0g||PgR, fmol/mg protein cytosol|
|pgr_cyt_pos|byte|%9.0g|janej|PgR cytosol>=25fmol/mg protein?|
|upa|float|%9.0g||uPA, ng/mg protein|
|upa_pos|byte|%9.0g|janej|uPA-positive? (Upper tertile)|
|her2_fish_pos|byte|%9.0g|janej|HER2+ with FISH?|
|ki67|byte|%9.0g||Ki67 - IHC-score|
|ki67_pos|byte|%9.0g|janej|Ki67>20%?|

Some comments to the list of variables:

- age --- Age at diagnosis, years
- ddfs --- Time in years from inclusion to diagnosis of first distant metastasis or to last follow-up if no metastases have been found.
- distmet --- Event indicator corresponding to the time variable ddfs.
- er_cyt --- Expression of the estrogen receptor alpha - a biomarker which is prognostic, and also, more importantly, predictive of the effect of tamoxifen treatment.
- pgr_cyt --- Expression of the progesteron receptor.
- upa --- Expression of the urokinase plasminogen activator---a factor related to metastatic spread.
- her2_fish_pos --- HER2-status. DNA amplification of the HER2 gene is a prognostic factor, but its role as a predictive marker for effect of Herceptin (trastuzumab) is more important.
- ki67 --- A proliferation marker measured using immunohistochemistry (IHC). It is used in clinical routine in Sweden as well as in many other countries. Chemotherapy is usually the choice for tumors with a high fraction of ki67-positive tumor cells nuclei. Scale: 0-100%.

The first report of the results from this study: [Malmström et al. J Clin Oncol. 2001 Apr 1;19(7):2010-9](https://doi.org/10.1200/jco.2001.19.7.2010)

### Assignment

```{r setup, message=FALSE}
# Load project data.
library(appstats2)

# Load Tidyverse packages for data manipulation.
library(tidyverse)

# Load gtsummary package for making summary tables.
library(gtsummary)

# Load survival and survminer for drawing Kaplan-Meier plots.
library(survival)
library(survminer)

# Load tidymodels to facilitate regression modelling.
library(tidymodels)
```

> The breast cancer data is loaded in this document using the command `library(appstats2)` in the code chunk above. For the sourcecode importing the data from the provided Excel file and formatting it as a tibble, please see the file [`breast_cancer.R`](https://github.com/JonasEngstrom/appstats2/blob/main/data-raw/breast_cancer.R). The Excel file [`breast_cancer.xls`](https://canvas.education.lu.se/courses/34580/files/6061926?module_item_id=1441783) referred to in the code can be found on Lund University’s website. To install the package `appstats2` run the command `devtools::install_github('JonasEngstrom/appstats2')` or read the file [`README.md`](https://github.com/JonasEngstrom/appstats2/blob/main/README.md) on [the package’s GitHub repo](https://github.com/JonasEngstrom/appstats2) for further instructions.

The focus of this assignment is on the prognostic value of the biomarker uPA.

1. Briefly describe the patient material:
    a. Some patient and tumor characteristics
    b. Treatment
    c. Time and status variable (ddfs and distmet)
    
    ```{r table-one}
    # Read the breast cancaer data set.
    breast_cancer |>
      
      # Relabel the values in the adjuvant variable, for better readability.
      mutate(adjuvant = if_else(adjuvant, 'Received', 'Not received')) |>
      
      # Construnct a summary table, i.e. a traditional “Table 1”.
      tbl_summary(
        
        # Choose what variables to include.
        include = c(
          age,
          ddfs,
          distmet,
          tumsize,
          hist_grade,
          mitotic_grade,
          tubular_grade,
          nuclear_grade,
          er_cyt,
          pgr_cyt,
          upa,
          her2_fish_pos,
          ki67
        ),
        
        # Choose what variable to group the table by.
        by = adjuvant,
        
        # Label the variables for better readability.
        label = list(
          age ~ 'Age',
          ddfs ~ 'Distant disease-free survival (years)',
          distmet ~ 'Distant metastases',
          tumsize ~ 'Tumor size (mm)',
          hist_grade ~ 'Histological grade',
          mitotic_grade ~ 'Mitotic grade',
          tubular_grade ~ 'Tubular grade',
          nuclear_grade ~ 'Nuclear grade',
          er_cyt ~ 'ER (fmol/mg)',
          pgr_cyt ~ 'PgR (fmol/mg)',
          upa ~ 'uPA (ng/mg)',
          her2_fish_pos ~ 'HER2 positivity',
          ki67 ~ 'Ki-67 IHC score'
        ),
        
        # Exclude missing values from the table, to reduce clutter.
        missing = 'no'
      ) |>
      
      # Label the grouping variable in the table header.
      modify_spanning_header(c('stat_1', 'stat_2') ~ '**Adjuvant treatment**') |>
      
      # Add totals for the variables in the table.
      add_overall() |>
      
      # Relocate the newly added Overall column to the left of the grouping variable.
      modify_table_body(~.x %>% relocate(stat_0, .after = label)) |>
      
      # Make the variable labels bold.
      bold_labels() |>
      
      # Add footnotes describing the numbers reported in the table.
      modify_footnote_header('Median (Q1, Q3); Frequency (%).', columns = all_stat_cols()) |>
      
      # Add footnote explaining abbreviations used in the table.
      modify_footnote_header('ER, estrogen receptor \U03B1; HER2; human epidermal growth factor receptor 2; IHC, immunohistochemistry; Ki-61, Kiel 67; PgR, progesteron receptor; uPA, urokinase plasminogen activator.', columns = 'label') |>
      
      # Add an overall caption to the table.
      modify_caption('Table 1. Patient characteristics by treatment received.')
    ```

2. Describe in more detail the characteristics of the groups with low and high uPA, separately (upa_pos=0 and upa_pos=1, respectively).

    ```{r table-two, message=FALSE}
    # Read the breast cancaer data set.
    breast_cancer |>
      
      # Relabel the values in the adjuvant and the upa_pos variables, for better readability.
      mutate(adjuvant = if_else(adjuvant, 'Received', 'Not received')) |>
      mutate(upa_pos = if_else(upa_pos, '< 0.231', '≥ 0.231')) |>
      
      # Construnct a summary table, i.e. a traditional “Table 1”.
      tbl_summary(
        include = c(
          age,
          ddfs,
          distmet,
          tumsize,
          hist_grade,
          mitotic_grade,
          tubular_grade,
          nuclear_grade,
          er_cyt,
          pgr_cyt,
          her2_fish_pos,
          ki67,
          adjuvant
        ),
        
        # Choose what variable to group the table by.
        by = upa_pos,
        
        # Choose what variables to include.
        label = list(
          age ~ 'Age',
          ddfs ~ 'Distant disease-free survival (years)',
          distmet ~ 'Distant metastases',
          tumsize ~ 'Tumor size (mm)',
          hist_grade ~ 'Histological grade',
          mitotic_grade ~ 'Mitotic grade',
          tubular_grade ~ 'Tubular grade',
          nuclear_grade ~ 'Nuclear grade',
          er_cyt ~ 'ER (fmol/mg)',
          pgr_cyt ~ 'PgR (fmol/mg)',
          her2_fish_pos ~ 'HER2 positivity',
          ki67 ~ 'Ki-67 IHC score',
          adjuvant ~ 'Adjuvant treatment'
        ),
        
        # Exclude missing values from the table, to reduce clutter.
        missing = 'no'
      ) |>
      
      # Label the grouping variable in the table header.
      modify_spanning_header(c('stat_1', 'stat_2') ~ '**uPA (ng/mg)**') |>
      
      # Add totals for the variables in the table.
      add_overall() |>
      
      # Relocate the newly added Overall column to the left of the grouping variable.
      modify_table_body(~.x %>% relocate(stat_0, .after = label)) |>
      
      # Make the variable labels bold.
      bold_labels() |>
      
      # Add footnotes describing the numbers reported in the table.
      modify_footnote_header('Median (Q1, Q3); Frequency (%).', columns = all_stat_cols()) |>
      
      # Add footnote explaining abbreviations used in the table.
      modify_footnote_header('ER, estrogen receptor \U03B1; HER2; human epidermal growth factor receptor 2; IHC, immunohistochemistry; Ki-61, Kiel 67; PgR, progesteron receptor; uPA, urokinase plasminogen activator.', columns = 'label') |>
      
      # Add an overall caption to the table.
      modify_caption('Table 2. Patient characteristics by uPA level.')
    ```

3. Describe the uPA distribution graphically.

    ```{r upa-linear-distribution, fig.cap='Figure 1: uPA distribution.', warning=FALSE, message=FALSE}
    # Load the breast cancer data set.
    breast_cancer |>
      
      # Make a plot with upa values on the x axis.
      ggplot(aes(x = upa)) +
      
        # Plot a histogram with blue, semitransparent bars.
        geom_histogram(
          fill = 'blue3',
          alpha = .5,
          color = 'blue3'
        ) +
      
        # Plot a horizontal boxplot along 80 on the y axis, with blue borders.
        geom_boxplot(
          aes(y = 80),
          width = 6,
          color = 'blue3'
        ) +
      
        # Plot individual observations along 80 on the y axis, represented by
        # blue semitransparent dots, spread slightly along the y axis, to
        # facilitate discrimination between the observations.
        geom_jitter(
          aes(y = 80),
          height = 1,
          alpha = .25,
          fill = 'blue3',
          alpha = .5,
          color = 'blue3'
        ) +
      
        # Change the figure labels, title, and caption.
        labs(
          x = 'uPA (ng/ml)',
          y = 'Count (for histogram)',
          title = 'uPA level distribution',
          caption = 'uPA; urokinase plasminogen activator.'
        ) +
      
        # Change the color theme of the graph.
        theme_minimal()
    ```

    ```{r upa-log-distribution, fig.cap='Figure 2: uPA distribution on logarithmic scale due to one extreme outlier.', warning=FALSE, message=FALSE}
    # Load the breast cancer data set.
    breast_cancer |>
      
      # Make a plot with upa values on the x axis.
      ggplot(aes(x = upa)) +
      
        # Plot a histogram with blue, semitransparent bars.
        geom_histogram(
          fill = 'blue3',
          alpha = .5,
          color = 'blue3'
        ) +
      
        # Plot a horizontal boxplot along 80 on the y axis, with blue borders.
        geom_boxplot(
          aes(y = 25),
          width = 3,
          alpha = .5,
          color = 'blue3'
        ) +
      
        # Plot individual observations along 25 on the y axis, represented by
        # blue semitransparent dots, spread slightly along the y axis, to
        # facilitate discrimination between the observations.
        geom_jitter(
          aes(y = 25),
          height = 1,
          alpha = .25,
          fill = 'blue3',
          alpha = .5,
          color = 'blue3'
        ) +
      
        # Convert the x scale to a logarithmic scale, to make the distribution
        # of observations clearer, since the data contains an extreme outlier.
        scale_x_log10() +
      
        # Change the figure labels, title, and caption.
        labs(
          x = 'uPA (log10(ng/ml))',
          y = 'Count (for histogram)',
          title = 'uPA level distribution',
          caption = 'uPA; urokinase plasminogen activator.'
        ) +
      
        # Change the color theme of the graph.
        theme_minimal()
    ```

4. Describe time to first distant metastasis graphically and estimate the fraction, in the underlying population, free from distant metastases five years after diagnosis of primary breast cancer.

    ```{r km-plot, fig.cap='Figure 3: Kaplan-Meier plot of time until distant metastases.'}
    # Create a survival curve from a survival object looking at distmet events
    # over ddfs time in the breast cancer dataset. Store the result in the
    # variable survival_fit.
    survival_fit <- survfit(Surv(ddfs, distmet) ~ 1, data = breast_cancer)
    
    # Calculate the proportion of patients who didn’t have a distmet event by
    # five years of follow up and store the result in the variable five_year.
    five_year <-
      survival_fit |>
      summary(times = 5)
    
    # Make a Kaplan-Meier plot and store it in the variable km_plot.
    km_plot <-
      survival_fit |>
      ggsurvplot(
        
        # Change axis labels and title of the plot.
        xlab = 'Follow-up time (years)',
        ylab = 'Proportion without distant metastases',
        title = 'Time until distant metastasis or last follow-up',
        
        # Set the line and confidence interval to be bloe and semitransparent.
        palette = alpha(c('blue3'), c(0.5))
      )
    
    # Plot the Kaplan-Meier plot.
    km_plot$plot +
      
      # Plot a dashed lines intersecting at the Kaplan-Meier plot at five
      # years of follow-up.
      annotate(
        'segment', x = 0, xend = 5, y = five_year$surv, yend = five_year$surv,
        linetype = 'dashed'
      ) +
      annotate(
        'segment', x = 5, xend = 5, y = 0, yend = five_year$surv,
        linetype = 'dashed'
      ) +
      
      # Change the color theme of the graph.
      theme_minimal() +
      
      # Remove the legend, as the plot only contains one group.
      theme(
        legend.position = 'none'
      ) +
      
      # Add the percentage without a distmet event after five years of
      # follow-up to the plot.
      annotate(
        "text",
        x = 5.5,
        y = .875,
        label = paste(round(five_year$surv * 100), '%')
      )
    ```

5. Analyze the association between uPA and metastasis-free survival. Describe it both numerically and graphically. Which measure is appropriate?

    ```{r km-plot-by-upa-pos, fig.cap='Figure 4: Kaplan-Meier plot of time until distant metastases by uPA level..'}
    # Create a survival curve from a survival object looking at distmet events
    # over ddfs time in the breast cancer dataset, stratified by the variable
    # upa_pos. Store the result in the  variable survival_fit.
    survival_fit_by_upa_pos <-
      survfit(Surv(ddfs, distmet) ~ upa_pos, data = breast_cancer)
    
    # Make a Kaplan-Meier plot and store it in the variable km_plot_by_upa_pos.
    km_plot_by_upa_pos <-
      survival_fit_by_upa_pos |>
      ggsurvplot(
        
        # Enable plotting of confidence intervals.
        conf.int = TRUE,
        
        # Change axis labels, title, and legend of plot.
        xlab = 'Follow-up time (years)',
        ylab = 'Proportion without distant metastases',
        title = 'Time until distant metastasis or last follow-up',
        legend.title = 'Urokinase plasminogen activator',
        legend.labs = c('< 0.231 ng/mg', '≥ 0.231 ng/mg'),
        
        # Change lines of plot to be blue and red and semitransparent.
        palette = alpha(c('blue3', 'red3'), c(0.5, 0.5))
      )
    
    # Plot the Kaplan-Meier plot.
    km_plot_by_upa_pos$plot +
      
      # Change color theme of plot.
      theme_minimal() +
      
      # Move the legend to the bottom of the plot.
      theme(
        legend.position = 'bottom'
      )
    ```

    ```{r log-model}
    # Train a logistic regression model using the locigal variable upa_pos
    # to predict dismet events.
    log_model <-
      glm(distmet ~ upa_pos, 'binomial', breast_cancer)
    
    # Show a summary of the trained model.
    log_model |>
      summary()
    ```

    ```{r log-model-odds-ratio}
    # Calculate the odds ratio for the second coefficient (i.e. ups_posTRUE)
    # of the model.
    log_model$coefficients[2] |>
      exp()
    ```

    ```{r upa-relative-risk, message=FALSE}
    # Calculate relative risk of ditmet events depending on ups_pos status.
    
    # Load the breast cancer dataset.
    breast_cancer |>
      
      # Group the data by ups_pos and distmet.
      group_by(upa_pos, distmet) |>
      
      # Calculate the total number of patients in each group combination.
      summarize(n = n()) |>
      
      # Ungroup the data.
      ungroup() |>
      
      # Drop missing data.
      drop_na() |>
      
      # Group by upa_pos.
      group_by(upa_pos) |>
      
      # Calculate the total for ups_pos TRUE and FALSE groups respectively.
      mutate(total = sum(n)) |>
      
      # Ungroup the data.
      ungroup() |>
      
      # Calculate the proportion of patients in each group.
      mutate(proportion = n / total) |>
      
      # Remove patients wihtout a distment event.
      filter(distmet) |>
      
      # Remove variables other than upa_pos and the proportion variable
      # previously calculate.
      select(upa_pos, proportion) |>
      
      # Pivot the data, so that patients with and without upa_pos TRUE are
      # reported in differnt variables.
      pivot_wider(names_from = upa_pos, values_from = proportion) |>
      
      # Calculate the proportion of patients with ups_pos TRUE to those with
      # FALSE, which gives the relative risk.
      mutate(relative_risk = `TRUE` / `FALSE`) |>
      
      # Extract the relative risk for output.
      pull(relative_risk)
    ```
    
    ```{r upa-grouped-box-pot, fig.cap='Figure 5: Presence of distant metastases by uPA level.', warning=FALSE}
    # Load the breast cancer dataset.
    breast_cancer |>
      
      # Relabel the variable distmet for better readability.
      mutate(distmet = if_else(distmet, 'Distant metastases', 'No distant metastases')) |>
      
      # Make a plot wiht distmit group membership on the x axis and upa level
      # on the y axis.
      ggplot(aes(x = distmet, y = upa)) +
      
        # Plot a grouped boxplot.
        geom_boxplot() +
      
        # Plot individual observations represented by semitransparent dots,
        # spread slightly along the x axis, to facilitate discrimination between
        # the observations, on top of the boxplots.
        geom_jitter(aes(color = distmet), width = .2, alpha = .4) +
      
        # Change axis labels, title, and caption of the plot.
        labs(
          y = 'uPA (log10(ng/mg))',
          title = 'Presence of distant metastases by uPA level',
          caption = 'uPA; urokinase plasminogen activator.'
        ) +
      
        # Set y scale to a logarithmic scale to increase clarity, as the data
        # contains one extreme outlier.
        scale_y_log10() +
      
        # Change the color theme of the plot.
        theme_minimal() +
      
        # Remove legend and x axis label from plot.
        theme(
          legend.position = 'none',
          axis.title.x = element_blank()
        ) +
      
        # Set dots to be read and blue.
        scale_color_manual(values = c('red3', 'blue3'))
    ```

6. Age is often a confounder in non-randomized studies. Evaluate if the effect of uPA is age dependent by adjusting for age in a multivariable Cox-model.

    ```{r cox-reg-age, fig.caption='Table 3: Cox regression.'}
    # Create a survival object looking at dismet events over ddfs time in the
    # breast cancer dataset and perform Cox regression by upa and age.
    cox_reg_age <-
      coxph(Surv(ddfs, distmet) ~ upa + age, data = breast_cancer)
    
    # Display a summary of the Cox regression.
    summary(cox_reg_age)
    
    # Display a table of the results of the Cox regression.
    tbl_regression(cox_reg_age, exponentiate = TRUE)
    ```

    ```{r cox-reg, fig.caption='Table 4: Cox regression.'}
    # Create a survival object looking at dismet events over ddfs time in the
    # breast cancer dataset and perform Cox regression by upa.
    cox_reg <-
      coxph(Surv(ddfs, distmet) ~ upa, data = breast_cancer)
    
    # Display a summary of the Cox regression.
    summary(cox_reg)
    
    # Display a table of the results of the Cox regression.
    tbl_regression(cox_reg, exponentiate = TRUE)
    ```

    ```{r age-groups}
    # Stratify patients in the breast cancer dataset by age group and store the
    # result in the variable age_groups.
    age_groups <-
      breast_cancer |>
      mutate(age_group = case_when(
        age < 35 ~ '< 35',
        age >= 35 & age < 40 ~ '35-40',
        age >= 40 & age < 45 ~ '40-45',
        age >=45 & age < 50 ~ '45-50',
        age > 50 ~ '> 50') |>
          ordered(levels = c(
            '< 35',
            '35-40',
            '40-45',
            '45-50',
            '> 50'
          )
        )
      ) |>
      
      # Remove the order from the factor levels in the age_group variable
      # and set the group > 50 as the reference group.
      mutate(
        age_group = factor(age_group, ordered=FALSE) |> relevel('> 50')
      )
    ```

    ```{r age-group-cox-reg, fig.caption='Table 5: Cox regression.'}
    # Create a survival object looking at dismet events over ddfs time in the
    # breast cancer dataset and perform Cox regression by upa and age stratum.
    cox_reg_age_groups <-
      coxph(Surv(ddfs, distmet) ~ upa + age_group, data = age_groups)
    
    # Display a summary of the Cox regression.
    summary(cox_reg_age_groups)
    
    # Display a table of the results of the Cox regression.
    tbl_regression(cox_reg_age_groups, exponentiate = TRUE)
    ```

    ```{r age-group-cox-reg-ups-strata, fig.caption='Table 6: Cox regression.'}
    # Create a survival object looking at dismet events over ddfs time in the
    # breast cancer dataset and perform Cox regression by upa_pos and age stratum.
    cox_reg_age_groups_ups_strata <-
      coxph(Surv(ddfs, distmet) ~ upa_pos + age_group, data = age_groups)
    
    # Display a summary of the Cox regression.
    summary(cox_reg_age_groups_ups_strata)
    
    # Display a table of the results of the Cox regression.
    tbl_regression(cox_reg_age_groups_ups_strata, exponentiate = TRUE)
    ```

7. Assume now instead that the aim is to build a prognostic model that can be used to stratify other patients in the underlying population according to risk of developing distant metastases. Which predictors would you like to include in the model and why and how (transformation)? This is not a trivial question. Too many parameters in the model will in general lead to overfit (modelling of noise) and hence bad performance when the model is validated in an independent material, but a model with too few parameters is no better since it will utilize the information in the sample poorly. This question has no correct answer. Argue for your choice of variables in the final model.

    ```{r prediction-model, fig.cap='Figure 6: Prediction model receiver operating characteristics curve.'}
    # Create a prediction model.
    
    # Make a pipeline to process the model’s input data.
    data_pipeline <-
      
      # Load the breast cancer dataset.
      breast_cancer |>
      
      # Initialize a pipeline, known as a recipe in R’s tidymodel package.
      recipe() |>
      
      # Chose independent variables.
      update_role(
        upa_pos,
        age,
        tubular_grade,
        nuclear_grade,
        mitotic_grade,
        tumsize,
        er_cyt,
        pgr_cyt,
        her2_fish_pos,
        new_role = 'predictor'
      ) |>
      
      # Chose dependent variable.
      update_role(
        distmet,
        new_role = 'outcome'
      ) |>
      
      # Recode the dependent variable as a factor.
      step_bin2factor(distmet, skip = TRUE) |>
      
      # Stratify patients by age.
      step_mutate(
        age_group = case_when(
          age < 35 ~ '< 35',
          age >= 35 & age < 40 ~ '35-40',
          age >= 40 & age < 45 ~ '40-45',
          age >=45 & age < 50 ~ '45-50',
          age > 50 ~ '> 50'
        ) |>
          factor()
      )
    
    # Train a logistic regression model.
    prediction_model <-
      
      # Initialize a workflow, the second part of tidymodel pipelines, that
      # takes a recipe (see above) and a model specification to train a model.
      workflow() |>
      
      # Define that we want to use a logistic regresion model.
      add_model(
        logistic_reg() |>
          set_engine('glm')
      ) |>
      
      # Add the recipe specified above.
      add_recipe(data_pipeline) |>
      
      # Fit the workflow model to the breast cancer dataset after
      # preprocessing the data using the recipe specified.
      fit(breast_cancer)
    
    # Store predictions and true values in the variable model_predictions in
    # preparation for evaluation of apparent model performance.
    model_predictions <-
      
      # Make predictions using the model.
      prediction_model |>
      predict(breast_cancer, type = 'prob') |>
      
      #Add the predictions to a column of the correct answers from input data.
      bind_cols(
        breast_cancer |>
          
          # Recode the correct answers as yes and no, to match the model
          # prediction output.
          mutate(
            truth = case_match(
              distmet,
              FALSE ~ 'no',
              TRUE ~ 'yes'
            ) |>
              factor() |>
              
              # Relevel the data so that yes is used as the reference. Another
              # option here would have been to use .pred_no in the following
              # statement calculating area under the ROC curve, but we chose
              # to do it this way.
              relevel('yes')
          ) |>
          select(truth)
      )
    
    # Calculate the area under the ROC curve for the model.
    auroc <- roc_auc(model_predictions, truth = truth, .pred_yes)
    
    # Print the area under the ROC curve.
    print(auroc)
    
    # Plot the ROC curve.
    roc_curve(model_predictions, truth = truth, .pred_yes) |>
      autoplot() +
      
      # Add the area under the curve as an annotation to the plot.
      annotate("text", x = .75, y = .25, label = paste('AUROC', round(auroc$.estimate, 4))) +
      
      # Add a title to the plot.
      labs(
        title = 'Receiver Operator Characteristics'
      ) +
      
      # Change the color theme of the graph.
      theme_minimal()
    ```

8. **Optional:** When including a covariate in a Cox-model, you implicitly assume that the hazard increases or decreases with the same multiplicative factor for each step on the scale of the variable.
    a. Fit a Cox-model with upa as the only covariate. How many percent does the hazard of distant metastases increase per step on the uPA-scale (per ng/mg protein)?
    b. Variables, like uPA, which are measured on a ratio scale, do often have a skewed distribution. In such cases, it is usually a good idea to log-transform the values. Create a new variable log2_upa = ln(upa)/ln(2), i.e. the logarithm with base 2, and fit a Cox-model model with log2_upa as the only covariate. How many percent does the hazard increase for each doubling of the uPA concentration?
    c. Is the fit of this model better than that of the model with uPA on a linear scale (without transformation)?
    d. How can the fit of two or more models be compared?
    e. Effects are seldom linear. Transformation of covariates is one way of taking nonlinearity into account; another way is categorization, but you should be aware of that you are throwing away potentially important prognostic information when categorizing variables measured on a continuous scale.
    f. A good rule is to select a simple model which explains a lot of the variability in the outcome variable. This may sound easy, but it is not. By fitting a model which is too complex you start modelling random noise and such a model will in general not perform well in validation studies, but also models which are too simple will validate badly. Fit a series of univariable Cox-models in which upa has been categorized into two groups of the same size (categorize at the median), into three groups of the same size (tertiles) and finally into four groups of the same size (quartiles). Define the upa-variables as categorical and select the category with the lowest values as the reference group. Interpret the estimated hazard ratios. For the model with upa in 4 groups: draw a forest plot of the HRs with 95% CIs relative to the reference category. Hint: Let the y-axis have 4 values (1-4) labelled Q1, Q2 vs Q1, Q3 vs Q1 and Q4 vs Q1, where Q is an abbreviation for Quartile. For each row, plot the HR vs Q1 on log-scale (base 2). For Q1 the HR vs Q1 will be deterministic 1.00 per definition (no 95% C). Since 1.00 corresponds to 0 on log2-scale (2^0^=1), plot it vs x=0 but label that x-value 1.0. For Q2 vs Q1 plot the corresponding HR and 95% CI and so on. The 95% CI:s will be symmetric around the HR-estimates on the log2-scale. Imagine, e.g., that a HR vs Q1 is 2.0, then the HR on log2-scale is 1.0 since 2^1^=2. In general plot the log2 HR, i.e. ln(HR)/ln(2) and transform the 95% CI-limits accordingly.

Good luck!
