---
title: "warm_up_exercises"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{warm_up_exercises}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Applied Statistics II—Warm-up exercises

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(appstats2)
library(tidyverse)
library(kableExtra)
```

The instructions below are copied from [the Lund University website](https://canvas.education.lu.se/courses/34580/files/6061874?module_item_id=1441692).

These computer exercises are meant to re-familiarize you with your statistical program and also to serve as a refresher on some of the topics covered during the course *Applied Statistics I*.

You will use the same dataset as in the previous course, SCANIA_SURVEY_1000. The dataset comprises 1000 observations that were randomly chosen from the data file Scania Survey (adults from the population of Skåne, 18 – 80 years old at baseline). The participants have provided answers on a wide range of survey questions about living conditions, life style, environmental aspects and health at three different occasions, year 2000, 2005 and 2010.

## 1. Table with descriptive statistics.

Create the variable BMI based on the participants height and weight in 2000.

```{r bmi-function-definition}
#' Body Mass Index
#' 
#' Calculated body mass index from height and weight.
#'
#' @param weight Body weight in kilograms.
#' @param height Body height in meters.
#'
#' @returns Body mass index in kilograms per square meters.
#' @export
#'
#' @examples
#' bmi(75, 1.81)
bmi <- function(weight, height) {
  weight / height ^ 2
}
```

```{r bmi-variable-creation}
bmi_table <-
  scania_survey_1000 |>
  mutate(bmi = bmi(weight_00, height_00 / 100))
```

Next create a table with descriptive statistics separately for smokers and non-smokers for the variables age, BMI and gender in 2000.

```{r descriptive-table}
bmi_table |>
  group_by(smoking_00, gender_00) |>
  summarize(
    min_age = min(age_00) |> round(1),
    mean_age = mean(age_00) |> round(1),
    sd_age = sd(age_00) |> round(1),
    max_age = max(age_00) |> round(1),
    min_bmi = min(bmi, na.rm = TRUE) |> round(1),
    mean_bmi = mean(bmi, na.rm = TRUE) |> round(1),
    sd_bmi = sd(bmi, na.rm = TRUE) |> round(1),
    max_bmi = max(bmi, na.rm = TRUE) |> round(1),
    n = n()
  ) |>
  ungroup() |>
  kbl() |>
  kable_classic()
```

## 2. Graphically examine the BMI distribution for smokers and non-smokers.

Does the distribution of BMI differ for smokers and non-smokers?

```{r bmi-distribution}
bmi_table |>
  mutate(smoking_00 = smoking_00 == 1) |>
  ggplot(aes(bmi, group = smoking_00, fill = smoking_00)) +
  geom_density(alpha = .5) +
  theme(legend.position = 'bottom')
```

## 3. Choose an appropriate test to examine if there is a significant BMI difference between smokers and non-smokers.

```{r bmi-test}
wilcox.test(bmi_table |> pull(bmi) ~ bmi_table |> pull(smoking_00))
```

a. which variable is the outcome and what is the type of the outcome?  
*BMI is the outcome. It is continuous.*

b. do you compare dependent or independent groups?  
*Independent groups.*

c. what is the conclusion?  
*There is a difference.*

## 4. Choose an appropriate test to examine if there is a significant association between gender and smoking.

```{r smoking-test}
chisq.test(scania_survey_1000 |> pull(smoking_00), scania_survey_1000 |> pull(gender_00))
```

a. which variable is the outcome and what is the type of the outcome?  
*Smoking is the outcome. It is dichotomous.*

b. what is the conclusion?  
*There is a difference.*
